Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Validation

### 2012-12-10 paper.tex
Updated some of the free-fall section.

A simple test to verify the Poisson solver implemented by CASTRO is
the case of gravitational free fall. In this setup, two stars, each
individually in an equilibrium state, are placed on the computational
grid, separated by an initial distance $r_0$ along the $x$ axis with
zero initial velocity. We choose stars of masses $0.6$ and $0.8\,
M_\odot$, with the lower mass star on the left side (e.g. $x < 0$) of
the domain and the higher mass star on the right, such that their
center of mass coincides with the center of the
domain. Gravitationally, the stars may be treated as point masses
until the point of contact, so the equation of motion governing the
distance $r$ between their centers of mass is that of simple free
fall:
\begin{equation}
  \ddot{r}(t) = - \frac{GM}{r},
\end{equation}
where $G$ is the gravitational constant and $M$ is the total mass of
the system. This differential equation has a closed-form solution for
the evolution time as a function of separation:
\begin{equation}
  t(r) = \sqrt{\frac{r_0^3}{2GM}} \left[ \text{arccos}\left(\sqrt{\frac{r}{r_0}}\,\right) + \sqrt{\frac{r}{r_0} \left(1 - \frac{r}{r_0}\right)}\ \right]. \label{analyticalFreeFall}
\end{equation} \MarginPar{is there a reference for this? or did you derive it?}
We determine the initial separation to be consistent with the
simulation performed in Section \ref{Sec:Kepler}; that is, we select
an initial orbital period and use Kepler's third law to calculate the
radius of a circular orbit with that period. We select an initial
orbital period of 200 s for this simulation, so that the separation is
\[
  r_0 = 5.73 \times 10^{9}\ \text{cm}.
\]
We chose a relatively low resolution simulation to demonstrate the
capabilities of CASTRO even while using modest resources. The
computational grid is covered by a coarse grid of $48^3$ zones, with
two levels of refinement above the coarse grid. Each refined grid
carries an increase in resolution by a factor of 4 relative to the
coarser grid below it. CASTRO initially assigns $93\%$ of the domain
to be covered by the intermediate resolution grids, and $0.05\%$ of
the domain to be covered by the finest resolution grids.

The analytical result in Equation \ref{analyticalFreeFall} determines
the total elapsed free-fall time,
\[
  t_{\text{ff}} = \frac{\pi}{2} \sqrt{\frac{r_0^3}{2GM}}.
\]
The physical radius of each white dwarf is roughly $10\%$ of the
initial separation, so we consider the evolution terminated when the
radial separation reaches that value (at that point, the stars will no
longer be in free-fall due to physical contact). Since the elapsed
time goes roughly as the square root of the distance for small $r$, we
consider the evolution terminated when $t = 0.99\, t_{\text{ff}}$. The
results of our simulation are shown in Figure \ref{Fig:Free Fall}.

>>>>

A simple test to verify the Poisson solver implemented by CASTRO is
the case of gravitational free fall. In this setup, two stars, each
individually in an equilibrium state, are placed on the computational
grid, separated by an initial distance $r_0$ along the $x$ axis with
zero initial velocity. We choose stars of masses $0.6$ and $0.8\,
M_\odot$, with the lower mass star on the left side (e.g. $x < 0$) of
the domain and the higher mass star on the right, such that their
center of mass coincides with the center of the
domain. Gravitationally, the stars may be treated as point masses
until the point of contact, so the equation of motion governing the
distance $r$ between their centers of mass is that of simple free
fall:
\begin{equation}
  \ddot{r}(t) = - \frac{GM}{r},
\end{equation}
where $G$ is the gravitational constant and $M$ is the total mass of
the system. This differential equation has a closed-form solution for
the evolution time as a function of separation:
\begin{equation}
  t(r) = \sqrt{\frac{r_0^3}{2GM}} \left[ \text{arccos}\left(\sqrt{\frac{r}{r_0}}\,\right) + \sqrt{\frac{r}{r_0} \left(1 - \frac{r}{r_0}\right)}\ \right]. \label{analyticalFreeFall}
\end{equation}\MarginPar{This result is derived in the freefall directory; I have edited the text to make it clear how I arrived at this result.}
This result can be derived by recognizing that
\[
  t(r) = \int_{r_0}^r \frac{dr}{v(r)}
\]
and inserting the velocity as a function of radial separation,
\[
  v(r) = \sqrt{\frac{2GM}{r_0}\left(\frac{r_0}{r} - 1\right)}.
\]
We determine the initial separation to be consistent with the
simulation performed in Section \ref{Sec:Kepler}; that is, we select
an initial orbital period and use Kepler's third law to calculate the
radius of a circular orbit with that period. We select an initial
orbital period of 200 s for this simulation, so that the separation is
\[
  r_0 = 5.73 \times 10^{9}\ \text{cm}.
\]
We chose a relatively low resolution simulation to demonstrate the
capabilities of CASTRO even while using modest resources. The
computational grid is covered by a coarse grid of $48^3$ zones, with
two levels of refinement above the coarse grid. Each refined grid
carries an increase in resolution by a factor of 4 relative to the
coarser grid below it. CASTRO initially assigns $93\%$ of the domain
to be covered by the intermediate resolution grids, and $0.05\%$ of
the domain to be covered by the finest resolution grids.

The analytical result in Equation \ref{analyticalFreeFall} determines
the total elapsed free-fall time,
\[
  t_{\text{ff}} = \frac{\pi}{2} \sqrt{\frac{r_0^3}{2GM}}.
\]
The physical radius of each white dwarf is roughly $10\%$ of the
initial separation, so we consider the evolution terminated when the
radial separation reaches that value (at that point, the stars will no
longer be in free-fall due to physical contact). Since the elapsed
time goes roughly as the square root of the distance for small $r$, we
consider the evolution terminated when $t = 0.99\, t_{\text{ff}}$. The
results of our simulation are shown in Figure \ref{Fig:Free Fall}. The
positions of the two stars are determined by calculating the center of
mass of the right $(x > 0)$ and left $(x < 0)$ sides of the domain at 
the end of each time step, and treating the centers of mass as the 
respective location of the two stars.
